---
title: "EDRM 718 Project A-Gun Deaths"
author: Hui Jiang
date: January 30 2021
output: html_notebook
---

### Description

This report will study the compositions of the intents of gun deaths and individual
attributes in American in 2012. The data for this study is primarily from the 
Centers for Disease COntrol ans Prevention's Multiple Cause of Death database 
provided by Ben Casselman, et al.
First, I will study the compositions of gun death types in the sample. I will 
then examine the proportions of different groups in terms of individual 
attributes.

*Source of data: Centers for Disease COntrol ans Prevention's Multiple Cause of Death database*

### Analysis

In this data set, there are four types of gun death: accidental, homicide, 
suicide, undetermined. Figure 1 illustrates the composition of the gun death 
types in 2012.

\newpage

***

*Figure 1. Bar plots with proportions of gun death types in 2012.*

```{r echo=FALSE}
# 2. Write a report to analyze the data from 2012. Specifically, we are
# interested in the proportions of different types of gun deaths, the 
# proportions of each gender type, and the proportion of each race type. 
# Your report should include graphs, descriptive statistics, and 
# accompanying narrative. You can see Instructor Report 1 for an example and
# watch the accompanying videos for an example of completing this type of 
# analysis. (Your report does not need a press release.)

# read the data and put it in an object called "gun"

gun <- read.csv( "full_data.csv", header = T, sep = ",")

# choose the data from 2012

gun_2012 <- subset(gun, gun$year == 2012)

# calculate the frequencies of different types of gun death in 2012

death_counts_2012 <- table(gun_2012$intent)

# calculate the proportions of different types of gun death in 2012

death_props_2012 <-prop.table(death_counts_2012)

# provide the bar plot of gun death proportions in 2012

barplot(death_props_2012,
        xlab = "Gun Death",
        ylab = "Proportions", 
        ylim = c(0, 0.8))

```

***

Table 1 shows the frequencies and percents for four types of gun death. The 
sample consists of 2% in the accidental death, 36% in the homicide, 62% in the 
suicide, and 1% in the undetermined death. Suicide is the most common type in 
these four types.

***

*Table 1. Frequencies and percents for each type of gun death in 2012.*

```{r echo=FALSE}
# put the frequencies and proportions in a table

round(cbind(death_counts_2012, death_props_2012), 2)

```

***

Figure 2 illustrates the composition of the gender types in gun death in 2012. 
From this figure, it is apparent that there are much more males than females.

\newpage

***

*Figure 2. Bar plots with proportions of gender types in 2012.*

```{r echo=FALSE}
# calculate the frequencies of different gender types in 2012

gender_counts_2012 <- table(gun_2012$sex)

# calculate the proportions of different gender types in 2012

gender_props_2012 <- prop.table(gender_counts_2012)

# provide the bar plot of the gender proportions in 2012

barplot(gender_props_2012,
        xlab = "Gender",
        ylab = "Proportions", 
        ylim = c(0, 1))

```

***

Table 2 shows the frequencies and percents for gender types. There are 14% 
females and 86% males, which are more than 6 times of females. 

***

*Table 2. Frequencies and percents for each gender in 2012.*

```{r echo=FALSE}

# put the frequencies and proportions in a table

round(cbind(gender_counts_2012, gender_props_2012), 2)

```

***

The sample divides race into Asian/Pacific Islander, black, Hispanic, 
native American/native Alaskan and white. As Figure 3 illustrated, white people 
have the highest gun death rate than other three race types. 

\newpage

***

*Figure 3. Bar plots with proportions of race types in 2012.*

```{r echo=FALSE}
# calculate the frequencies of different race types in 2012

race_counts_2012 <- table(gun_2012$race)

# calculate the proportions of different race types in 2012

race_props_2012 <- prop.table(race_counts_2012)
race_props_2012

# provide the bar plot of the race proportions in 2012

barplot(race_props_2012,
        xlab = "Race",
        ylab = "Proportions", 
        ylim = c(0, 0.8))

```

***

Table 3 shows the frequencies and percents for race types. There are 1% in 
Asian/Pacific islander, 24% in black, 9% in Hispanic, 1% in native 
American/native Alaskan and 65% in white.  

***

*Table 3. Frequencies and percents for each race in 2012.*

```{r echo=FALSE}

# put the frequencies and proportions in a table

round(cbind(race_counts_2012, race_props_2012), 2)

```

***

### Conclusions

I examined the compositions of gun deaths types in American in 2012 and knew the
proportions of different groups in terms of gender and race. The most primary 
intent of gun deaths is suicide and then followed by homicide. The proportion 
of accidental deaths and undetermined deaths are only 2% and 1% respectively.

As for the individual attributes, there are more males in gun deaths than 
females. Specifically, 86 percent are males and 14 percent are females. White 
people have a larger proportions to die from gunshot than other races, i.e., 
black people, Hispanic, Asian/Pacific islander and native American/native 
Alaskan in turn. The results are helpful to take measures to decrease the gun 
deaths rate for different groups.

***

```{r}
# 6. Add to your notebook an analysis of the 2019 data by addressing the same 
# questions you addressed for 2012.

# read the 2019 data and put it in an object called "gun_2019" 
gun_2019 <- read.csv( "full_data_19.csv", header = T, sep = ",")

# calculate the frequencies of different types of gun death in 2019

death_counts_2019 <- table(gun_2019$intent)

# calculate the proportions of different types of gun death in 2019

death_props_2019 <-prop.table(death_counts_2019)

death_props_2019

# provide the bar plot of gun death proportions in 2019

barplot(death_props_2019,
        xlab = "Gun Death",
        ylab = "Proportions", 
        ylim = c(0, 0.8))

# calculate the frequencies of different gender types in 2019

gender_counts_2019 <- table(gun_2019$sex)

# calculate the proportions of different gender types in 2019

gender_props_2019 <- prop.table(gender_counts_2019)
gender_props_2019

# provide the bar plot of the gender proportions in 2019

barplot(gender_props_2019,
        xlab = "Gender",
        ylab = "Proportions", 
        ylim = c(0, 1))

# calculate the frequencies of different race types in 2019

race_counts_2019 <- table(gun_2019$race)

# calculate the proportions of different race types

race_props_2019 <- prop.table(race_counts_2019)
race_props_2019

# provide the bar plot of the race proportions in 2019

barplot(race_props_2019,
        xlab = "Race",
        ylab = "Proportions", 
        ylim = c(0, 0.8))

```


```{r}

# 7. For each of the variables of interest, compare proportions for the two 
# years (2012 and 2019) with graphs and descriptive statistics.

# combine the full_data.csv and full_data_19.csv as a new file full_data_all.csv 
# and read it

gun_all <- read.csv("full_data_all", header = T, sep = ",")


# choose the data from 2012 and 2019

gun_201219 <- subset(gun, gun$year == 2012 || 2019)

# construct a two-way table of frequencies of gun deaths type based on year

ses_death_counts <- table(gun_201219$intent, gun_201219$year)

#construct conditional percentages of gun deaths type

ses_death_prop <- 100 * prop.table(ses_death_counts, 2)
round(ses_death_prop)

# provide a stacked bar plot

barplot(ses_death_prop,
        xlab = "Year",
        ylab = "Percents of gun deaths",
        legend = c("accidental", "homicide", "suicide", "undetermined"),
        args.legend = list(title = "Gun deaths type", cex = .7))

```

```{r}

# construct a two-way table of frequencies of gender type based on year

ses_gender_counts <- table(gun_201219$sex, gun_201219$year)

#construct conditional percentages of gun deaths type

ses_gender_prop <- 100 * prop.table(ses_gender_counts, 2)
round(ses_gender_prop)

# provide a stacked bar plot

barplot(ses_gender_prop,
        xlab = "Year",
        ylab = "Percents of gender",
        legend = c("Female", "Male"),
        args.legend = list(title = "Gender", cex = .7))
```

```{r}

# construct a two-way table of frequencies of race type based on year

ses_race_counts <- table(gun_201219$race, gun_201219$year)

#construct conditional percentages of race type type

ses_race_prop <- 100 * prop.table(ses_race_counts, 2)
round(ses_race_prop)

# provide a stacked bar plot

barplot(ses_race_prop,
        xlab = "Year",
        ylab = "Percents of race",
        legend = c("Asian/Pacific Islander", "Black", "Hispanic", "Native American/Native Alaskan", "White"),
        args.legend = list(title = "Race", cex = .7))

```

